The Strucplot Framework:\\ Visualizing Multi-way Contingency Tables with \pkg{vcd}

### 2005-10-25 trunk/vcd/paper/JSS.Rnw
small typo
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@486 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
KH changes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@489 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% an abstract and keywords
\Abstract{
This paper describes the `strucplot' framework for the visualization of
multi-way contingency tables. Strucplot displays include hierarchical
conditional plots such as mosaic, association, and sieve plots, and
can be combined into more complex, specialized plots for visualizing
conditional independence, GLMs, and the results of
independence tests. The frameworks' modular design allows
flexible customization of the plots' graphical appearance, including 
shading, labeling, spacing, and legend, by means of \underline{gr}aphical
\underline{ap}pearance \underline{con}trol (`grapcon') functions. 
The framework is provided by the \proglang{R} package \pkg{vcd}.
}

@@

In order to explain multi-dimensional categorical data, statisticians
typically look for (conditional) independence structures. Whether the
task is purely exploratory or model-based, techniques such as mosaic and
association plots offer good support for visualization. Both have been
extended over the last two decades, and implementations exist in many
statistical environments. 
Both graphical methods visualize aspects of (possibly higher-dimensional)
contingency tables. A \emph{mosaic plot} \citep{vcd:Hartigan+Kleiner:1984} is basically 
an area-proportional visualization of (typically, observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. An \emph{association plot} \citep{vcd:Cohen:1980}
visualizes the standardized deviations of observed frequencies from
those expected under a certain independence hypothesis.
Each cell is represented by a rectangle that has
(signed) height proportional to the residual and width proportional
to the square root of the expected counts, 
so that the area of the box is proportional
to the difference in observed and expected frequencies.

Over the years, extensions to these techniques were mainly focused on
five aspects:

@@

Figures \ref{fig:arthritis} to \ref{fig:titanic} illustrate some of these
extensions (the code is shown in Appendix \ref{sec:code}). 
Figure \ref{fig:arthritis} shows the results from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, using an
extended mosaic plot with residual-based shading based on the
maximum statistic: clearly, the new treatment is effective. Figure
\ref{fig:ucbadmissions} visualizes the well-known UCB admissions data
by the means of a conditional association plot. The panels show the
residuals from a conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men accepted than
would be expected under the null) causes the rejection of the
hypothesis of conditional independence. Figure \ref{fig:presex}
illustrates the conditional independence of premarital and
extramarital sex, given gender and marital status. The Chi-squared
test of independence rejects the null hypothesis: possibly, because
too many men who had premarital sex also had extramarital sex?
Finally, Figure \ref{fig:titanic} visualizes the `Survival on the Titanic' 
data using a double-decker plot. Here, a binary response (survival to the disaster)
shall be explained by other factors (class, gender, and age). The blue
boxes represent the proportion of survived passengers in a particular
stratum. The proportions of saved women and children are indeed higher
than those of men, but they clearly decrease from the 1st to the 3rd class. In
addition, the proportion of saved men in the 1st class is higher than
in the others.

@@

This paper describes the strucplot framework provided by the \pkg{vcd} package
for the \proglang{R} system and environment for statistical computing
and graphics, available from the Comprehensive \proglang{R} Archive Network (\url{http://cran.r-project.org/}).
The framework integrates displays such as mosaic, association, and
sieve plots by their unifying property of being 
flat representations of contingency tables.
These basic plots, as well as specialized displays for conditional
independence, can be used both for exploratory visualization and
model-based analysis. Exploratory techniques include specialized displays for
the bivariate case, as well as pairs and trellis plot-like displays for
higher-dimensional tables. Model-based tools include 
methods suitable for the visualization of 
conditional independence tests (including permutation tests), as well
as for the visualization of particular GLMs (logistic regression,
log-linear models). Additionally, two of the frameworks' further strengths are its
flexibility and extensibility: graphical appearance aspects such as shading,
labeling, and spacing are modularized by means of `graphical
appearance functions', allowing fine-granular customization and
user-level extensions.

@@

Spacing of strucplot displays is customizable in a similar way than shading. The
\code{spacing} argument of the \codefun{strucplot} function takes a
list of \class{unit} vectors, one for each dimension, specifying the
space between the tiles corresponding to the levels. As an example,
consider the introductory example of the \data{Arthritis} data 
(Figure \ref{fig:arthritis}). Since we are interested in the effect of
the medicament in the placebo and treatment groups, a mosaic plot is
certainly appropriate to visualize the three levels of `Improved' in
the two `Treatment' strata. Another conceptual approach is to use
spine plots with highlighting \citep{vcd:hummel:1996}. A spine plot is
a variation of a bar plot where the heights of the bars are held
constant, whereas the widths are used to represent the number of cases
in each category. This is equivalent to a mosaic plot for a one-way table.
If a second (indicator) variable is highlighted in a spine plot, we
obtain a display equivalent to a simple mosaic display for a two-way
table, except that no space between the levels of the highlighted
variable is used.
In the \data{Arthritis} example, we will highlight patients with `Marked' 
improvement in both groups. To obtain such a display within 
the strucplot framework\footnote{The \pkg{vcd} package also provides 
a specialized \codefun{spine} function for spine plots and spinograms.}, it
suffices to set the space between the `Improved' tiles to 0 (see
Figure \ref{fig:artspine}):

@@

In this paper, we describe the `strucplot' framework for the visualization of
multi-way contingency tables. Strucplot displays include popular 
basic plots such as mosaic, association, and sieve plots, integrated 
in a unified framework: all can be seen as graphical visualizations of
hierarchical conditional flat tables. 
Additionally, these core strucplot displays can be combined into more complex, 
specialized plots, such as pairs and trellis-like displays for visualizing
conditional independence. Residual-based shadings permit the
visualization of log-linear models and the results of independence tests.
The frameworks' modular design allows
flexible customization of the plots' graphical appearance, including 
shading, labeling, spacing, and legend, by means of graphical
appearance control (`grapcon') functions. These `graphical
hyperparameters' are customized and created by
generating functions. Our work includes a set of predefined grapcon generators for
typical analysis tasks, and user-level extensions can easily be added.

>>>>

%% an abstract and keywords
\Abstract{
This paper describes the `strucplot' framework for the visualization of
multi-way contingency tables. Strucplot displays include hierarchical
conditional plots such as mosaic, association, and sieve plots, and
can be combined into more complex, specialized plots for visualizing
conditional independence, GLMs, and the results of
independence tests. The framework's modular design allows
flexible customization of the plots' graphical appearance, including 
shading, labeling, spacing, and legend, by means of \underline{gr}aphical
\underline{ap}pearance \underline{con}trol (`grapcon') functions. 
The framework is provided by the \proglang{R} package \pkg{vcd}.
}

@@

In order to explain multi-dimensional categorical data, statisticians
typically look for (conditional) independence structures. Whether the
task is purely exploratory or model-based, techniques such as mosaic and
association plots offer good support for visualization.  Both visualize
aspects of (possibly higher-dimensional) contingency tables, with
several extensions introduced over the last two decades, and
implementations available in many statistical environments.  A
\emph{mosaic plot} \citep{vcd:Hartigan+Kleiner:1984} is basically an
area-proportional visualization of (typically, observed) frequencies,
composed of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. An \emph{association plot}
\citep{vcd:Cohen:1980} visualizes the standardized deviations of
observed frequencies from those expected under a certain independence
hypothesis.  Each cell is represented by a rectangle that has (signed)
height proportional to the residual and width proportional to the square
root of the expected counts, so that the area of the box is proportional
to the difference in observed and expected frequencies.

Extensions to these techniques have mainly focused on the following
aspects.

@@

Figures \ref{fig:arthritis} to \ref{fig:titanic} illustrate some of these
extensions (the code is given in Appendix \ref{sec:code}). 
Figure \ref{fig:arthritis} shows the results from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, using an
extended mosaic plot with residual-based shading based on the
maximum statistic: clearly, the new treatment is effective. Figure
\ref{fig:ucbadmissions} visualizes the well-known UCB admissions data
by the means of a conditional association plot. The panels show the
residuals from a conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men accepted than
would be expected under the null) causes the rejection of the
hypothesis of conditional independence. Figure \ref{fig:presex}
illustrates the conditional independence of premarital and
extramarital sex, given gender and marital status. The Chi-squared
test of independence rejects the null hypothesis: possibly, because
too many men who had premarital sex also had extramarital sex?
Finally, Figure \ref{fig:titanic} visualizes the `Survival on the Titanic' 
data using a double-decker plot. Here, a binary response (survival to the disaster)
shall be explained by other factors (class, gender, and age). The blue
boxes represent the proportion of survived passengers in a particular
stratum. The proportions of saved women and children are indeed higher
than those of men, but they clearly decrease from the 1st to the 3rd class. In
addition, the proportion of saved men in the 1st class is higher than
in the others.

@@

This paper describes the strucplot framework provided by the \pkg{vcd}
package for the \proglang{R} environment for statistical computing and
graphics, available from the Comprehensive \proglang{R} Archive Network
(\url{http://cran.r-project.org/}).  The framework integrates displays
such as mosaic, association, and sieve plots by their unifying property
of being flat representations of contingency tables.  These basic plots,
as well as specialized displays for conditional independence, can be
used both for exploratory visualization and model-based analysis.
Exploratory techniques include specialized displays for the bivariate
case, as well as pairs and trellis plot-like displays for
higher-dimensional tables. Model-based tools include methods suitable
for the visualization of conditional independence tests (including
permutation tests), as well as for the visualization of particular GLMs
(logistic regression, log-linear models). Additionally, two of the
framework's further strengths are its flexibility and extensibility:
graphical appearance aspects such as shading, labeling, and spacing are
modularized by means of `graphical appearance functions', allowing
fine-granular customization and user-level extensions.

@@

Spacing of strucplot displays is customizable in a similar way than shading. The
\code{spacing} argument of the \codefun{strucplot} function takes a
list of \class{unit} vectors, one for each dimension, specifying the
space between the tiles corresponding to the levels. 
Consider again the introductory example of the \data{Arthritis} data 
(Figure \ref{fig:arthritis}). Since we are interested in the effect of
the medicament in the placebo and treatment groups, a mosaic plot is
certainly appropriate to visualize the three levels of `Improved' in
the two `Treatment' strata. Another conceptual approach is to use
spine plots with highlighting \citep{vcd:hummel:1996}. A spine plot is
a variation of a bar plot where the heights of the bars are held
constant, whereas the widths are used to represent the number of cases
in each category. This is equivalent to a mosaic plot for a one-way table.
If a second (indicator) variable is highlighted in a spine plot, we
obtain a display equivalent to a simple mosaic display for a two-way
table, except that no space between the levels of the highlighted
variable is used.
In the \data{Arthritis} example, we will highlight patients with `Marked' 
improvement in both groups. To obtain such a display within 
the strucplot framework\footnote{The \pkg{vcd} package also provides 
a specialized \codefun{spine} function for spine plots and spinograms.}, it
suffices to set the space between the `Improved' tiles to 0 (see
Figure \ref{fig:artspine}):

@@

In this paper, we describe the `strucplot' framework for the visualization of
multi-way contingency tables. Strucplot displays include popular 
basic plots such as mosaic, association, and sieve plots, integrated 
in a unified framework: all can be seen as graphical visualizations of
hierarchical conditional flat tables. 
Additionally, these core strucplot displays can be combined into more complex, 
specialized plots, such as pairs and trellis-like displays for visualizing
conditional independence. Residual-based shadings permit the
visualization of log-linear models and the results of independence tests.
The framework's modular design allows
flexible customization of the plots' graphical appearance, including 
shading, labeling, spacing, and legend, by means of graphical
appearance control (`grapcon') functions. These `graphical
hyperparameters' are customized and created by
generating functions. Our work includes a set of predefined grapcon generators for
typical analysis tasks, and user-level extensions can easily be added.
